Dissertation Proposal

### 2023-11-01 proposal.tex
clean up RQ1

\usepackage[utf8]{inputenc} % Use UTF-8 encoding
\usepackage[T1]{fontenc} % Use the T1 font encoding
\usepackage{geometry} % Adjust page margins
\usepackage{amsmath, amssymb} % For math symbols and equations
\usepackage{graphicx} % For including images
\usepackage{titlesec} % For customizing section titles
\usepackage[style=authoryear]{biblatex}  % For managing references
\addbibresource{bibfile.bib}
\usepackage{hyperref}
\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\miscon}[1]{\textbf{#1}}
\newcommand{\surmisedMiscon}[1]{\miscon{#1}$^\ddag$}
\newcommand{\pValue}[0]{\textit{p}\nobreakdash-value}

@@

>>>>

\usepackage[utf8]{inputenc} % Use UTF-8 encoding
\usepackage[T1]{fontenc} % Use the T1 font encoding
\usepackage{geometry} % Adjust page margins
\usepackage{amsmath, amssymb} % For math symbols and equations
\usepackage{graphicx} % For including images
\usepackage{titlesec} % For customizing section titles
\usepackage[style=authoryear]{biblatex}  % For managing references
\addbibresource{bibfile.bib}
\usepackage{hyperref}
\usepackage[colorinlistoftodos]{todonotes}
\newcommand{\miscon}[1]{\textbf{#1}}
\newcommand{\surmisedMiscon}[1]{\miscon{#1}$^\ddag$}
\newcommand{\pValue}[0]{\textit{p}\nobreakdash-value}
\newcommand{\rqI}{What are the program-behavior misconceptions that apply to most modern programming languages and present (even) in students with prior programming backgrounds?}
\newcommand{\rqII}{What would be a tutoring system that aims at correcting the misconceptions, draws on cognitive and educational psychology concepts, utilizes a notional machine, and seems to be effective?}

@@

\section{Background: Misconceptions}%
\label{s:background-misconceptions}

\section{Background: Program Behavior}

\section{Plan for \textbf{RQ1}}%
\label{s:plan-rq1}

\begin{description}
  \item[RQ1] \rqI
\end{description}

To answer this RQ, I need to first define what I mean by ``apply to
most modern programming languages''. As I described in the
Introduction (\Cref{s:intro}), most modern languages share a semantic
core. I have modelled this core as SMoL (\Cref{s:smol}). Because SMoL
represents the semantic core, misconceptions about SMoL presumably
also apply to most modern PLs.

To find misconceptions that ``present (even) in students with
prior programming backgrounds?'', I have been collecting data from
multiple populations that very likely have prior experience. For
example, my priliminary results are build upon data collected in a
upper-year programming language course at Brown.

To find a reasonably comprehensive collection of misconceptions, I
seek for inputs from students. Most prior works
(\Cref{s:background-misconceptions}) have experts coming up with a
list of potential misconceptions and then filter the potential
misconceptions with student inputs. This approach can't possibly find
misconceptions not considered by experts and hence is necessarily
affected by experts' blind spots. To overcome experts' blind spots, I
analyze a dataset produced by a two-year process to find (a)
student-written programs that tend to trip up students and (b)
incorrect responses to those programs (\Cref{s:gen-prob-quizius}),
using a tool for the purpose (\Cref{s:quizius}). I have been cleaning
up those programs (\Cref{s:smol-quizzes,s:tutor-refine-program-set})
and confirming that these programs are still tricky in the same way
(\Cref{s:misconceptions-quiz,s:misconceptions-tutor}).

I have been creating a tutoring system to answer my RQ2. The tutoring
system collects more student data, which becomes another source of
misconceptions. The tutoring system, which I explain in details when I
talk about RQ2 (\Cref{s:plan-rq2}), includes many tasks that asks
students to predict the result(s) of running a program. I plan to
change the tutor such that it asks students' for their explanations
when they give a wrong answer. I plan to eyeball the explanations to
see if there are any new misconceptions.

To be reasonably specific about misconceptions, I have created a
definitional interpreter for each misconception that I have found. I
plan to do the same things for potentially newly found/confirmed
misconceptions.

Eventually, I expect to build a table of misconceptions. Each
misconception comes with two or more programs. For each of these
programs, the wrong result that correspond to the misconception should
be commonly predicted by students from various population.
Furthermore, these students give similar explanations for all the
wrong answers. For example, I might find a misconception called
\miscon{DefByRef}: when students are asked to predict the result of
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
X\% of them responded \lstinline|12 0|; and when asked to predict
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! x 0)
  x
  y
\end{lstlisting}
Y\% of them responded \lstinline|0 12|; and these students all
explained something like ``\lstinline|y| is \lstinline|x|''.
Preliminary results can be found in \Cref{s:misconceptions-tutor}.

\section{Plan for \textbf{RQ2}}%
\label{s:plan-rq2}

\begin{description}
  \item[RQ2] \rqII
\end{description}

Prior works suggest that presenting refutation text and encouraging
students to compare cases are helpful for learning, and that notional
machines are useful for explaning program behavior.

I would first explain the major task format
(\Cref{s:interpreting-task}) in the tutor and then explain how I would
let the tutor incorporate ideas of refutation text
(\Cref{s:refutationt-text}), case comparsion
(\Cref{s:case-comparison}), and notional machines
(\Cref{s:notional-machine}). Finally, I explain how I plan to evaluate
the effectiveness of the tutor (\Cref{s:evaluation}).

\subsection{Interpreting Task}%
\label{s:interpreting-task}

To build a tutor that incorporate these ideas, I plan to includes lots
of multiple-choice questions (MCQs) in the tutor. I call these
interpreting tasks because they ask students to interpret programs:
each MCQ presents a program, for example,
\begin{lstlisting}
  (defvar x 12)
  (defvar y x)
  (set! y 0)
  x
  y
\end{lstlisting}
and then asks students to choose the results of the program.

\paragraph{The MCQ Options} The options include the correct answers and
wrong answers that correspond to known misconceptions (in this case,
only \lstinline|0 0|, which correspond to \miscon{DefByRef}), among
other options that I will explain shortly. Because students might give
wrong answers that I do not anticipate, I plan to include an ``Other''
option that, once choosen, allows students to enter arbitrary answer.
If an MCQs only have a small number of options, students have a great
chance (in this case, 50\%) of just guessing the right answer or
successfully using a process of elimination. So I plan to add
additional options that are not obviously wrong: options generated by
moving and/or copying numbers in the current options. In this case, I
will add \lstinline|12 12| and \lstinline|0 12|. In summary, each MCQ
has the following options:
\begin{itemize}
  \item the correct answer,
  \item wrong answers that correspond to known misconceptions,
  \item options generated by moving and/or copying numbers in the
  current options
  \item the ``Other'' option (which, once choosen, allows students to
  enter arbitrary answer)
\end{itemize}

\paragraph{Panelty to Wrong Choices} To discourage students from
making a random choice, the tutor should gives panelty. Deciding the
panelty is a non-trivial question. A common strategy is to grade
students by their correctness. However, grading by correctness is not
suitable for a tutoring system because we do not mind students doing
badly on earlier tasks as long as they eventually master the concepts.
Actually, if students generally do perfectly on all tasks, that is a
failure of the tutoring system -- it is likely not teaching anything
to the students. If grading by correctness is not a suitable option, a
common alternative is to grade by completion. However, grading by
completion alone encourage students to rush through the tutor and
hence might encourage students to make random choices. In short, I
would like to make students feel as follows:
\begin{itemize}
  \item They won't lose grade as long as they finish the tutor.
  \item They should avoid giving wrong answers.
\end{itemize}
To encourage this feeling, I do the following: when I deployed the
tutor in my institution, I graded by completion\footnote{However, I
have no control over how other institutions use the tutor.}; when a
student fails a task, the tutor gives an \emph{extra} task:
\begin{enumerate}

  \item The program is semantically the same as the first, but with
    superficial changes (e.g., variable names, constants, and
    operators are changed) so that the student cannot immediately
    guess the answer.

  \item Instead of multiple-choice, students must \emph{type} the
    answer into a text box. (The tutor normalizes text to accommodate
    variations.) This is intentional. First, I want to force
    reflection on the explanation just given, whereas with an MCQ,
    students could just guess. Second, I feel that students would
    find typing more onerous than clicking. In case students had just
    guessed on a question, my hope is that the penalty of having to
    type means, on \emph{subsequent} tasks, they would be more likely
    to pause and reflect before choosing.

\subsection{Design the Tutor with Refutation Text and Other Explanations}%
\label{s:refutationt-text}

The tutor presents a refutation text when a student seems to hold a
misconception. Let's reconsider the program in
\Cref{s:interpreting-task}. If a student answers \lstinline|0 0|,
which suggests they hold the \miscon{DefByRef} misconception, the
tutor responds
\begin{quote}
  The correct answer is \lstinline|12 0|. \lstinline|y| was bound to
  \lstinline|12| (i.e., the value of \lstinline|x|) rather than to
  \lstinline|x|. So changing the value of \lstinline|y| does not
  changes the value of \lstinline|x|.
\end{quote}
If a students gave a wrong answer that does not correspond to a
misconception (either by choosing one of the generated option or by
choosing the ``Other'' option), the tutor presents a generic
explanation. For example, if a student answers \lstinline|12 12|, the
tutor responds
\begin{quote}
  The correct answer is \lstinline|12 0|. The first definition binds
  \lstinline|x| to \lstinline|12|. The second definition binds
  \lstinline|y| to the value of \lstinline|x|, which is
  \lstinline|12|. The \lstinline|set!| mutates the binding of
  \lstinline|y|, so \lstinline|y| is now bound to \lstinline|0|.
\end{quote}

\subsection{Design the Tutor with Case Comparison}%
\label{s:case-comparison}

The tutor provides considerable opportunities for making case
comparison: each interpreting task presents a SMoL program and the
program's result. These program-result pairs are concrete cases of the
semantics of SMoL. To incorporate the idea of case comparsion, the
tutor prompts students to compare these cases and then to summarize
the rules of program behavior.

It is impractical to ask students to compare tens or even hundreads of
programs. So

\subsection{Design the Tutor with Case Comparison}%
\label{s:notional-machine}

\subsection{Evaluate the Tutor}%
\label{s:evaluation}
